Is Partner Modeling Mutual?

### 2014-03-20 project-mutual-modeling.tex
Formatting, syntax

\documentclass{report}

@@

@@

@@

\begin{document}

\title{Mutual Modeling in Human-Robot Interaction - Scientific part}

\chapter{Scientific Part}
\section{Summary}

Human-robot interaction is a challenge for artificial intelligence. This field
indeed lays at the crossroad of several other domains of AI and requires to
tackle them in a holistic manner: Modeling humans and human cognition;
acquiring, representing, manipulating in a tractable way abstract knowledge;
reasoning on this knowledge to make decisions; and eventually instantiating
those decisions into physical actions in coordination with humans. Many AI
techniques are invited, from visual processing to symbolic reasoning, from task
planning to \emph{theory of mind} building, from reactive control to action
recognition and learning.

\paragraph{Theory of Mind}

Theory of Mind (originally defined in~\cite{premack1978does}) is the cognitive
ability that a subject possesses to represent the mental state of another
agent, possibly including knowledge that contradicts the subject's own model: for
example, a book can be at the same time \emph{visible} for myself, and \emph{not
visible} for you.

Children develop this skill, which is essential to understand others'
perspectives during interactions, around the age of three. It supposes the
ability to build, store and retrieve separate models of the knowledge of the
interactors.

One classical application of this cognitive skill is the so-called
\emph{False-Belief} experiment (also known as the \emph{Sally and Ann}
experiment)~\cite{Leslie2000}: a child is asked to watch a scene where two
people, A and B, manipulate objects. Then A leaves and B hides away one
object. When A comes back, we ask the child ``where do you think A will
look for the object?''. Before acquiring a theory of mind, children are not
able to separate their own (true) model of the world (where they know that
the object was hidden) from the model of A, which contains \emph{false
beliefs} on the world (A still thinks the object is at its original
position since he did not see B hiding it).

\paragraph{Mutual Modelling}

@@

A partner model is probably not a "box", \ie not a monolithical representation
but rather a mosaic of information fragments about the partner, with various
grain sizes and various life cycles. This mosaic is elaborated through various
mechanisms, first for building an initial model of the partner and then for
updating this model.  As two students meet for the first time, mutual models are
initialized by the assumptions they make upon each other from cues such as
his/her membership to large categories (age, culture, profession, ...) include
stereotypes (sportsmen, junkie, business women, Swiss,...) as well as physical
appearance. Scholars studied how initial modeling impacts communication. In
their experiments on initial mutual modelling, Slugoski, Lalljee, Lamb \&
Ginsburg~\cite{slugoski1993attribution} pretended to their subjects that their
(fake) partner had or had not received the same information. They observed that
the subjects adapted their dialogue by focusing the explanation on the items
that (s)he was supposed to ignore. Brennan (1991) showed that the subjects used
different initial strategies in forming queries depending on who they were told
their partner was.

@@

\section{Research Plan}

\subsection{State of the Art}

\paragraph{Building a model of a human during the interaction} Perspective
Taking is a human ability which allows one to put him/herself in another
person's point of view. Studied in psychology
literature~\cite{Flavell1992,Tversky1999}, this ability is crucial when
interacting with people by allowing one to reason on others' understanding of
the world in terms of visual perception, spatial descriptions, affordances and
beliefs, etc.  In the last years these notions have been gradually employed in
HRI.~\cite{Breazeal2006} presents a learning algorithm that takes into account
information about a teacher's visual perspective in order to learn a task.
~\cite{Johnson2005} apply visual perspective taking for action recognition
between two robots.~\cite{Trafton2005} use both visual and spatial perspective
taking to find out the referent indicated by a human partner.

The cognitive model that the robot builds for the agent it interacts with is
still simple and mostly focused on geometric features (who sees what? What are
our relative positions? etc.). Extending this knowledge with more subtle
perceptions (emotional state for instance) remains to be done.

\paragraph{Joint action}
Several theories dealing with collaboration~\cite{Cohen1991,Grosz1996,Clark1996}
emphasize that collaborative tasks have specific requirements compared to
individual ones, \eg, since the robot and the person share a common goal, they
have to agree on the manner to realize it, they must show their commitment to
the goal during execution, etc. Several robotic systems have already been built
based on these theories~\cite{Rich1997,Sidner2005,Breazeal2003} and they all
have shown benefits of this approach. They have also shown how difficult it is
to manage turn-taking between communication partners and to interleave task
realization and communication in a generic way. Finally, today only few
systems~\cite{Fong2006,Breazeal2003,Sisbot2008} take humans into account at all
levels.

\subsection{Related research of the applicants}

Pierre Dillenbourg has been working since several years on the role and importance
of mutual modelling in human-human interaction, specifically in the context of
learning sciences.

He already conducted several experiments~\cite{sangin2007partner,
nova2007collaboration} that supported the idea that modelling partners is not only
an individual skill. When engaged in a social task that could benefit from
mutual modelling, the accuracy of the models build by each of the participant on
the other participants increase.

In the frame of this work, Pierre Dillenbourg has proposed a simple formalism
that we describe below, and that we will use subsequently to describe the mutual
modelling situations.

\paragraph{A simple notational system for mutual modeling} We suggest
the notation \model{A}{B}{X} to denote ``A knows that B knows X". This is a
simple but hopefully useful reduction to reason on mutual modeling. This
notation does not mean A has an explicit, monolithic representation of B: it
must be understood as an abstraction referring to complex socio-cognitive
processes. Besides, we refer to the \emph{degree of accuracy} of the model as
\Model{A}{B}{X}.

To estimate the mutual modelling effort, we need 3 variables:

@@

    \item Tasks vary a lot with respect to how much they require.  The grounding
        criterion -- denoted $\mathcal{M}^{\circ}_{min}$ -- refers to how
        important it is to mutually share a piece of information X to succeed
        the task T. It can be computed as the probability to succeed T despite
        the fact X is not grounded. $\mathcal{M}^{\circ}_{min}(A,B,X)$ can be
        estimated from the correlation between \Model{A}{B}{X} and task
        performance.

@@

The experiments we report here address MM across different tasks, some with
dyads, other with triads. They were conducted along 6 years in two different
institutions by different researchers. They used different independent,
intermediate and dependent variables. Nonetheless, we were able to address a few
questions across these studies. We start with 3 simple hypotheses about
$\mathcal{M}^{\circ}(A,B)$:

@@

\subsection{Research agenda}

\subsection{Importance of the work}

\subsection{References}

@@

\section{International collaboration}

>>>>

\documentclass{article}

@@

\usepackage[draft]{fixme}

@@

\newcommand{\groundingcriterion}{{$\mathcal{M}^{\circ}_{min}$}}
\newcommand{\inigrounding}{{$\mathcal{M}^{\circ}_{t_0}$}}

@@

\title{Is Partner Modeling Mutual?}

\begin{document}
\maketitle

\begin{abstract}
    TDB
\end{abstract}

\section{Introduction}

From its very beginning, CSCL research has been following Roschelle \& Teasley's
(1986) suggestion that collaborative learning has something to do with the
process of constructing and maintaining a shared understanding of the task at
hand. Building a shared/mutual understanding refers to the upper class of
collaborative learning situations, those in which students should build upon
each other's understanding to refine their own understanding. What is expected
to produce learning is not the mere fact that two students build the same
understanding but the cognitive effort they have to engage to build this shared
understanding (Schwartz, 1995). This effort can be observed by the frequency of
rich interactions, i.e. interactions whose occurrence has been related to
learning: (self-) explanations in cognitive science (Chi ; Webb), conflict
resolution in socio-cognitive theories (Doise), mutual regulation (Blaye,) in a
Vygostkian perspective, etc. The construction of a shared understanding has been
investigated for several years in psycholinguistics, under the  notion of
“grounding” (Clark \&Wilkes-Gibbs, 1986). However, the relevance of grounding
mechanisms for explaining learning outcomes has been questioned. The monitoring
and repair of mis-understanding explains for instance referential failures in
short dialogue episodes but does hardly predict conceptual change over longer
sessions (Dillenbourg \& Traum, 2006). The cumulative effect of grounding
episodes can probably be better understood from a socio-cultural perspective:
"collaborative learning is associated with the increased cognitive-interactional
effort involved in the transition from learning to understand each other to
learning to understand the meanings of the semiotic tools that constitute the
mediators of interpersonal interaction" (Baker, Hansen, Joiner \& Traum, 1999).
Several scholars suggest that CSCL research should go deeper in the
understanding of how partners engage into shared meaning making (Stahl, 2006) or
'intersubjective' meaning making (Suthers (2006).

Paradoxically, while Clark's theory is somewhat too linguistic from a conceptual
change viewpoint, it is criticized at the same time as being too cognitivist by
some psycholinguists, i.e. as overestimating the amount shared knowledge and
mutual representations actually necessary to conduct a dialogue. The fundamental
issue, as old as philosophy, is the degree of coupling between the different
levels of dialogue, mostly between the lexical / syntactical level and the
deeper semantic levels. Pickering \& Garrod (2004) argue that the mutual
understanding starts mostly with a 'local alignment' at the level of the
linguistic representations, due to priming mechanisms, and that this superficial
alignment may – in some cases- lead to a 'global alignment' of the semantic
level.  For these authors, the convergence in dialogue, and even the repair of
some misunderstandings, is explained by this mimetic behavior more than by a
monitoring of each other's knowledge: "…interlocutors do not need to monitor and
develop full common ground as a regular, constant part of routine conversation,
as it would be unnecessary and far too costly. Establishment of full common
ground is, we argue, a specialized and non-automatic process that is used
primarily in times of difficulty (when radical misalignment becomes apparent)."
(p. 179). This view is actually not incompatible with Clark's grounding
criterion (REF): the degree of shared understanding that peers need to reach
depends upon the task they perform. For instance, a dialogue between two
surgeons might rely on superficial alignment if they talk about their friends
but has to guarantee accurate common grounds when talking about which
intervention will be conducted in which way on which patient.  Some CSCL
scholars stressed the 'illusion of shared understanding' (JACCO?), i.e. the
potential decoupling between linguistic alignment and actual shared meanings.

This interesting cognitive science debate mostly occurred outside the field of
learning. In education, the question is to relate these mechanisms to learning
outcomes. Is linguistic alignment sufficient to trigger conceptual change? Does
negotiation of meaning only occurs when partners monitor and diagnose each
other's knowledge. If the ratio between shallow alignment and deep grounding
depends upon the task, and if deep grounding is a condition for learning, then
the pedagogical challenge is to design tasks that require deep grounding. Most
empirical studies on grounding and alignment are conducted with tasks, which
despite being qualified as 'ecologically valid' by their authors, are mere
referencing tasks such as asking the way to the train station or helping the
peer to choose a picture among many. In this contribution, we explore several
richer tasks such as arguing on a sensitive issue or building a concept map.

Deep grounding or shared meaning making requires some cognitive load. For Clark
\& Wilkes-Gibbs (1986), what is important is not the individual effort made by
the receiver of a communicative act, but the overall least collaborative effort.
The cost of producing a perfect utterance may be higher than the cost of
repairing the problems that may arise through misunderstandings. For instance,
subjects are less careful about adapting their utterances to their partner when
they know they can provide feedback on his/her understanding (Schober, 1993). We
introduced the notion of ‘optimal collaborative effort’ (Dillenbourg et al,
1996) to stress that misunderstanding should not be viewed as something to be
avoided (if this was possible), but as an opportunity to engage into
verbalization, explanation, negotiation, and so forth. This issue is related to
the global argument regarding cognitive load in learning activities, especially
in discovery learning environments: there is no learning without some cognitive
load, but overload may hinder learning (Paas, Renkl \& Sweller, 2003). In the
context of collaborative learning, we understand the cognitive load induced by
mutual modeling as part of Schwartz’s (1995) notion of effort towards a shared
understanding. For instance, CSCL researchers expanded the use of 'collaboration
scripts' (Aronson,?). A script is a pedagogical method that collaborative
learning activities in order to foster the emergence of productive interactions
such as argumentation, explanation or conflict. Conflict-resolution scripts such
as the ArgueGraph (Dillenbourg \& Hong, 2008) forms pair of students with
opposite opinions, which increases the difficulty of consensus building,
requiring more justifications, more negotiation, more load. Similarly, JIGSAW
scripts provide peers with different but complementary knowledge for augmenting
(reasonably) the effort group members have to engage to reach a shared solution.

%\paragraph{Theory of Mind}
%
%Theory of Mind (originally defined in~\cite{premack1978does}) is the cognitive
%ability that a subject possesses to represent the mental state of another
%agent, possibly including knowledge that contradicts the subject's own model: for
%example, a book can be at the same time \emph{visible} for myself, and \emph{not
%visible} for you.
%
%Children develop this skill, which is essential to understand others'
%perspectives during interactions, around the age of three. It supposes the
%ability to build, store and retrieve separate models of the knowledge of the
%interactors.
%
%One classical application of this cognitive skill is the so-called
%\emph{False-Belief} experiment (also known as the \emph{Sally and Ann}
%experiment)~\cite{Leslie2000}: a child is asked to watch a scene where two
%people, A and B, manipulate objects. Then A leaves and B hides away one
%object. When A comes back, we ask the child ``where do you think A will
%look for the object?''. Before acquiring a theory of mind, children are not
%able to separate their own (true) model of the world (where they know that
%the object was hidden) from the model of A, which contains \emph{false
%beliefs} on the world (A still thinks the object is at its original
%position since he did not see B hiding it).

\section{Mutual Modeling}

@@

A partner model is probably not a "box", \ie not a monolithical representation
but rather a mosaic of information fragments about the partner, with various
grain sizes and various life cycles. This mosaic is elaborated through various
mechanisms, first for building an initial model of the partner and then for
updating this model.  As two students meet for the first time, mutual models are
initialized by the assumptions they make upon each other from cues such as
his/her membership to large categories (age, culture, profession, ...) include
stereotypes (sportsmen, junkie, business women, Swiss,...) as well as physical
appearance. Scholars studied how initial modeling impacts communication. In
their experiments on initial mutual modelling, Slugoski, Lalljee, Lamb \&
Ginsburg~\cite{slugoski1993attribution} pretended to their subjects that their
(fake) partner had or had not received the same information. They observed that
the subjects adapted their dialogue by focusing the explanation on the items
that (s)he was supposed to ignore. Brennan (1991) \fixme{ref} showed that the subjects used
different initial strategies in forming queries depending on who they were told
their partner was.

@@

%\subsection{State of the Art}
%
%\paragraph{Building a model of a human during the interaction} Perspective
%Taking is a human ability which allows one to put him/herself in another
%person's point of view. Studied in psychology
%literature~\cite{Flavell1992,Tversky1999}, this ability is crucial when
%interacting with people by allowing one to reason on others' understanding of
%the world in terms of visual perception, spatial descriptions, affordances and
%beliefs, etc.  In the last years these notions have been gradually employed in
%HRI.~\cite{Breazeal2006} presents a learning algorithm that takes into account
%information about a teacher's visual perspective in order to learn a task.
%~\cite{Johnson2005} apply visual perspective taking for action recognition
%between two robots.~\cite{Trafton2005} use both visual and spatial perspective
%taking to find out the referent indicated by a human partner.
%
%The cognitive model that the robot builds for the agent it interacts with is
%still simple and mostly focused on geometric features (who sees what? What are
%our relative positions? etc.). Extending this knowledge with more subtle
%perceptions (emotional state for instance) remains to be done.
%
%\paragraph{Joint action}
%Several theories dealing with collaboration~\cite{Cohen1991,Grosz1996,Clark1996}
%emphasize that collaborative tasks have specific requirements compared to
%individual ones, \eg, since the robot and the person share a common goal, they
%have to agree on the manner to realize it, they must show their commitment to
%the goal during execution, etc. Several robotic systems have already been built
%based on these theories~\cite{Rich1997,Sidner2005,Breazeal2003} and they all
%have shown benefits of this approach. They have also shown how difficult it is
%to manage turn-taking between communication partners and to interleave task
%realization and communication in a generic way. Finally, today only few
%systems~\cite{Fong2006,Breazeal2003,Sisbot2008} take humans into account at all
%levels.
%
\section{A model for mutual modelling evaluation}

To report experiments on mutual modeling, we suggest
the notation \model{A}{B}{X} to denote ``A knows that B knows X". This is a
simple but useful reduction to reason on mutual modelling. This
notation does not mean A has an explicit, monolithic representation of B: it
must be understood as an abstraction referring to complex socio-cognitive
processes. Besides, we refer to the \emph{degree of accuracy} of the model as
\Model{A}{B}{X}.

We parametrize and assess the mutual modelling effort through 3 variables:

@@

    \item Tasks vary a lot with respect to how much they require.  The grounding
        criterion -- denoted \groundingcriterion -- refers to how
        important it is to mutually share a piece of information X to succeed
        the task T. It can be computed as the probability to succeed T despite
        the fact X is not grounded. $\mathcal{M}^{\circ}_{min}(A,B,X)$ can be
        estimated from the correlation between \Model{A}{B}{X} and task
        performance.

@@

The experiments we report here address mutual modelling across different tasks, some with
dyads, other with triads. They were conducted along 6 years in two different
institutions by different researchers. They used different independent,
intermediate and dependent variables. Nonetheless, we were able to address a few
questions across these studies. We start with 3 simple hypotheses about
$\mathcal{M}^{\circ}(A,B)$:

@@

@@
